---
title: "selection_coefficient"
author: "WANG Danmeng"
date: "2026-05-07"
output: html_document
---
# set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  
                      message = FALSE,
                      warning = FALSE)

setwd("C:/Users/danme/Desktop/stage/code/selection_coef/block1")
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lmerTest)
library(ggh4x)
library(readxl)
library(ggsignif)
```

# data

```{r}
dat <- read.csv("predicted_frequencies_block1.csv", sep = ";")
pop_info <- read_xlsx("pop_information.xlsx")

sample <- dat %>%
  filter(is.na(calib))%>%
  dplyr::select(pop,mix, pop_id, generation, predicted_freq) %>%
  left_join(pop_info[,c(1,4)], by = "pop")

```

## data reformat

```{r}
# all data for plot
setup_map <- data.frame(
  pop_id = rep(LETTERS[1:8], each = 4),
  pop_id_new = unlist(lapply(LETTERS[1:8], function(x) {
    paste0(x, sprintf("%02d", 1:4))
  }))
)

setup_expanded_all <- sample %>%
  filter(generation == 0) %>%
  left_join(setup_map, by = "pop_id") %>%
  mutate(pop_id = pop_id_new) %>%
  dplyr::select(-pop_id_new) %>%
  mutate(id = substr(pop_id,2,3))%>%
  mutate(experiment_batch = case_when(id %in% c("01","02") ~ "RA",
                                      TRUE ~ "DW"))%>%
  dplyr::select(-id)
non_setup_all <- sample %>%
  filter(generation != 0)
all_data <- bind_rows(non_setup_all, setup_expanded_all) %>%
  arrange(pop_id, generation)

# mean data
mean_data <- all_data %>%
  group_by(pop_id, generation, pop, mix, experiment_batch) %>%
  dplyr::summarise(
    mean_freq = mean(predicted_freq, na.rm = TRUE)
  )

```

```{r, fig.width=6, fig.height=4}
mix_colors <- c(
  A = "#4E79A7",
  B = "#F28E2B",
  C = "#59A14F",
  D = "#E15759",
  E = "#B07AA1",
  F = "#76B7B2",
  G = "#EDC948",
  H = "#9C755F"
)

ggplot(mean_data, aes(x = generation, y = mean_freq, group = pop_id, color = mix)) +
  geom_line(alpha = 0.5,linewidth = 1.2) +
  scale_color_manual(values = mix_colors)+
  scale_x_continuous(
    breaks = c(0, 1, 4, 8)) +
  scale_y_continuous(
    limits = c(0.4, 0.8)) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("Estimated rec-1 mutant frequency")
```


```{r}
ggplot() +
  geom_point(data = all_data, 
             aes(x = generation, y = predicted_freq, group = pop_id, color = mix),
             alpha = 0.2)+
  geom_line(data = mean_data, 
            aes(x = generation, y = mean_freq, group = pop_id, color = mix),
            alpha = 0.4, linewidth = 1.5) +
  scale_color_manual(values = mix_colors)+
  scale_x_continuous(breaks = c(0, 1, 4, 8)) +
  scale_y_continuous(limits = c(0.4, 0.8)) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("Estimated rec-1 mutant frequency")
```


# selection coefficient


## logit

```{r}
logit_data <- all_data %>%
  mutate(logit_freq = qlogis(predicted_freq))
logit_mean <- mean_data %>%
  mutate(logit_mean_freq = qlogis(mean_freq))

ggplot() +
  geom_point(data = logit_data, 
             aes(x = generation, y = logit_freq, group = pop_id, color = mix),
             alpha = 0.2)+
  geom_line(data = logit_mean, 
            aes(x = generation, y = logit_mean_freq, group = pop_id, color = mix),
            alpha = 0.4, linewidth = 1.5) +
  scale_color_manual(values = mix_colors)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit rec-1 mutant frequency")
```


## check data distribution

```{r}
hist(x = mean_data$mean_freq, breaks = 30)
hist(x = logit_mean$logit_mean_freq, breaks = 30)
```

## fit model

### fit complexe

fixed effect : intercept and generation

random effect (on intercept and slope): mix and population within mix

-> warning: Model failed to converge with max|grad| = 0.00403952 (tol = 0.002, component 1)

-> meaning that model is too complexe (random effect structure)

-> corr for pop_id:mix btw intercept and generation is -1

```{r}
fit_complexe <- lmer(
  logit_mean_freq ~ 1+ generation + (generation | mix / pop_id ),
  data = logit_mean
)

summary(fit_complexe)
```

### fit simplified - good

fixed effect : intercept + generation

random effect on intercept : mix

random  effect on slope : mix and pop within mix

```{r}
fit_simple <- lmer(
  logit_mean_freq ~ 1+ generation + (0+generation| mix/pop_id ) + (1 | mix),
  data = logit_mean
)

summary(fit_simple)
```

### fit batch effect

fixed effect : intercept + generation

random effect on intercept : mix + batch

random  effect on slope : mix and pop within mix

Anova P value >> 0.05 --> no significant influence of batch effect

--> should use fit_simple

```{r}
fit_batch <- lmer(
  logit_mean_freq ~ 1+ generation + (0+generation| mix/pop_id ) + (1 | mix) +
                    (1 | experiment_batch),
  data = logit_mean
)

summary(fit_batch)
anova(fit_batch, fit_simple)
```

### fit binomial

```{r}
fit_binomial <- glmer(
  mean_freq ~ 1+ generation + (0+generation| mix/pop_id ) + (1 | mix),
  data = logit_mean,
  family = "binomial"
)

summary(fit_binomial)
```



#### significance

CI larger than 0 and P less than 0.05, so selection coefficient is significantly positive.
```{r}
# confidence interval
ci_s <- confint(fit_simple,parm = "generation",method = "Wald")
```



#### population level slope

```{r}
# population level slope deviation
# extract the modes of the random effects = population slope - overall slope
pop_re <- ranef(fit_simple)$`pop_id:mix` # extract the modes of the random effects
# extract random effect of mix
mix_re <- ranef(fit_simple)$mix %>%
  dplyr::select(generation) %>%
  tibble::rownames_to_column("mix") %>%
  dplyr::rename(mix_random_effect = generation)
# overall slope
s_overall <- fixef(fit_simple)["generation"]
# compute population slope
pop_slopes <- pop_re %>%
  tibble::rownames_to_column("pop_mix") %>%
  mutate(pop = substr(pop_mix, 1,3)) %>%
  mutate(mix = substr(pop_mix, 5,5))%>%
  dplyr::rename(pop_random_effect = generation) %>%
  left_join(mix_re, by = "mix") %>%
  mutate(
    selection_coef = s_overall + pop_random_effect + mix_random_effect)%>%
  left_join(pop_info[, c(3,4)], by = c("pop" = "pop_id"))
```



# Visualization

## fig1 general slope

```{r}
# generate regression line data
pred_grid <- data.frame(
  generation = seq(0,8,length.out = 200))

X <- model.matrix(~ generation, data = pred_grid) # design matrix
# compute y
# y = x * beta 
pred_grid$fit <- as.numeric(X %*% fixef(fit_simple))
# standard error of the predicted value
# Var(y) = X * Var(beta) * X^T
pred_grid$se <- sqrt(diag(X %*% vcov(fit_simple) %*% t(X)))
# boundary
pred_grid$lwr <- pred_grid$fit - 1.96 * pred_grid$se
pred_grid$upr <- pred_grid$fit + 1.96 * pred_grid$se
```


```{r}
ggplot() +
  geom_ribbon(data = pred_grid,
              aes(x = generation, ymin = lwr, ymax = upr),
              inherit.aes = FALSE, fill = "grey", alpha = 0.6) +
  geom_point(data = logit_data, 
             aes(x = generation, y = logit_freq, group = pop_id, color = mix),
             alpha = 0.3)+
  geom_line(data = logit_mean, 
            aes(x = generation, y = logit_mean_freq, group = pop_id, color = mix),
            alpha = 0.3, linewidth = 1.5) +
  geom_line(data = pred_grid, aes(x = generation, y = fit),
            inherit.aes = FALSE, color = "black", linewidth = 1.2) +
  scale_color_manual(values = mix_colors)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("Logit rec-1 mutant frequency")
```
```{r, fig.width=10, fig.height=4}
p_obs <- ggplot() +
  geom_point(data = logit_data,aes(x = generation,y = logit_freq,
                                   group = pop_id,color = mix),alpha = 0.3) +
  geom_line(data = logit_mean,aes(x = generation,y = logit_mean_freq,
                                  group = pop_id,color = mix),
            alpha = 0.3,linewidth = 1.5) +
  scale_color_manual(values = mix_colors) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12)) +
  xlab("Generation") +
  ylab("Logit rec-1 mutant frequency")

p_pred <- ggplot() +
  geom_ribbon(data = pred_grid,
              aes(x = generation,ymin = lwr,ymax = upr),fill = "grey",alpha = 0.6) +
  geom_line(data = pred_grid,aes(x = generation,y = fit),
            color = "black",linewidth = 1.2 ) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12)) +
  xlab("Generation") +
  ylab("Predicted logit rec-1 mutant frequency")


y_range <- range(
  logit_data$logit_freq,
  logit_mean$logit_mean_freq,
  pred_grid$lwr,
  pred_grid$upr,
  na.rm = TRUE
)
p_obs <- p_obs + coord_cartesian(ylim = y_range)
p_pred <- p_pred + coord_cartesian(ylim = y_range)

library(patchwork)

p_obs + p_pred
```



## fig2 population slope

```{r}
mix_levels <- levels(as.factor(pop_slopes$mix))
strip_cols <- mix_colors[seq_along(mix_levels)]

ggplot( pop_slopes, aes(x = pop, y = selection_coef)) +
  geom_hline( yintercept = s_overall, linetype = "dashed", color = "grey40", 
              linewidth = 1.2) +
  geom_segment( aes(x = pop, xend = pop,
                    y = selection_coef, yend = s_overall,color = mix),
                linetype = "dashed", linewidth = 1, alpha = 0.7) +
  geom_point( aes(color = mix), size = 3, alpha = 0.7) +
  scale_color_manual(values = mix_colors)+
  theme_classic() +
  facet_nested(~ mix, scales = "free_x", space = "free_x",
               strip =strip_themed( background_x = elem_list_rect(
               fill = scales::alpha(strip_cols, 0.7), 
               colour = strip_cols)))+
  xlab("Population") +
  ylab("Selection coefficient") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 9),
    axis.text.x = element_text( angle = 45, hjust = 1, vjust = 1
    )
  )
```

## fig3 population slope distribution

```{r, fig.height=3.5, fig.width=4}
# set position of overall point
dens <- density(pop_slopes$selection_coef, na.rm = TRUE)
y_s <- max(dens$y) * 0.15
ci_df <- data.frame(
  xmin = ci_s[1],
  xmax = ci_s[2],
  y = y_s,
  x = s_overall
)

ggplot(pop_slopes, aes(x = selection_coef)) +
  geom_density( fill = "grey", alpha = 0.5, color = "black", linewidth = 1) +
  geom_vline( xintercept = s_overall, linetype = "dashed", linewidth = 1) +
  geom_errorbar(data = ci_df, aes(y = y, xmin = xmin, xmax = xmax),inherit.aes = FALSE,
                orientation = "y", width = 2, linewidth = 1) +
  geom_point( aes( x = s_overall, y = y_s), 
              inherit.aes = FALSE, size = 3 ) +
  theme_classic() +
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Selection coefficient") +
  ylab("Density")
```

# Comparaison with Tom's data

## selection coefficient

### load data

```{r}
load("direct_selection_scoef_TOM.Rdata") # scoef data frame

# filter data of strain EEV1401, initial freq = 0.5
TOM_selection <- scoef %>%
  filter(genotype == "EEV1401" & initialfreq_aimed == "0.5" ) %>%
  rename(pop = id, selection_coef = s) %>%
  mutate(experiment_batch = "TP")%>%
  mutate(env = case_when(env == "nacl" ~ "Salt",
                         TRUE ~ "NGM"))

# create new dat for all selection coefficient
all_scoef <- pop_slopes %>%
  dplyr::select(pop, selection_coef, experiment_batch) %>%
  mutate(env = "Zinc") %>%
  bind_rows(TOM_selection[,c(1,2,8,9)])%>%
  filter(!is.na(selection_coef))
```




```{r}
ggplot(data = all_scoef, aes(x= env, y = selection_coef, color = env))+
  geom_boxplot()+
  geom_jitter(width = 0.2, size = 3, alpha = 0.5)+
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Environment") +
  ylab("Selection coefficient")
```

```{r, fig.width=10, fig.height=4}
all_scoef_summary <- all_scoef %>%
  group_by(env) %>%
  summarise(
    n = sum(!is.na(selection_coef)),
    mean_scoef = mean(selection_coef, na.rm = TRUE),
    sd_scoef = sd(selection_coef, na.rm = TRUE),
    se_scoef = sd_scoef / sqrt(n),
    ci95 = qt(0.975, df = n - 1) * se_scoef,
    .groups = "drop"
  )

ggplot(all_scoef_summary, aes(x = env, y = mean_scoef, color = env)) +
  geom_point(size = 4) +
  geom_hline( yintercept = 0, linetype = "dashed", color = "grey80", 
              linewidth = 1.2) +
  geom_errorbar(
    aes(
      ymin = mean_scoef - ci95,
      ymax = mean_scoef + ci95
    ),
    width = 0.2,
    linewidth = 0.8
  ) +
  scale_y_continuous(breaks = seq(-0.1,0.1, by = 0.05))+
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  theme_classic() +
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12)
  ) +
  xlab("Environment") +
  ylab("Selection coefficient")
```


