---
title: "selection_coefficient"
author: "WANG Danmeng"
date: "2026-05-07"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---
# set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  
                      message = FALSE,
                      warning = FALSE)

setwd("C:/Users/danme/Desktop/stage/code/selection_coef/block1")
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lmerTest)
library(ggh4x)
library(readxl)
library(ggsignif)
library(stringr)
```

# data

```{r}
dat <- read.csv("predicted_frequencies_block2.csv", sep = ";") # 653 obs
pop_info <- read_xlsx("pop_information.xlsx")
```

## data reformat


```{r}
sample <- dat %>%
  filter(is.na(calib))%>% # 473obs
  filter(!is.na(predicted_freq)) %>%# 413 obs
  dplyr::select(sample,pop, setup, generation, competition, predicted_freq, experiment_batch, plate_qpcr_id)%>%
  mutate(comp_back = case_when(competition == "V5"~"ECA722",
                               competition == "V6"~"ECA191",
                               competition == "V7"~"ECA730",
                               competition == "V8"~"ECA723"))
```


```{r}
# all data for plot
setup_map <- pop_info[1:48,c(1,4)]

setup <- sample %>%
  filter(generation == 0) %>%
  left_join(setup_map, by = "setup")%>%
  dplyr::select(-pop.x)%>%
  dplyr::rename("pop"="pop.y")%>%
  dplyr::relocate(pop, .before = 1)

non_setup <- sample %>%
  filter(generation != 0)

all_data <- bind_rows(non_setup, setup) %>%
  arrange(pop, generation)%>%
  mutate( env = case_when( str_detect(pop, "Z") ~ "Zinc",
                           TRUE ~ "NGM"))

# mean data
mean_data <- all_data %>%
  group_by(pop, generation, competition, comp_back, experiment_batch, env) %>%
  dplyr::summarise(
    mean_freq = mean(predicted_freq, na.rm = TRUE)
  )

```

```{r, fig.width=10, fig.height=3}
ggplot(mean_data, aes(x = generation, y = mean_freq, group = pop, color = env)) +
  facet_wrap(~comp_back, nrow = 1)+
  geom_line(alpha = 0.5,linewidth = 1.2) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("rec-1 mutant frequency")
```


```{r}
ggplot() +
  facet_wrap(~comp_back, nrow = 2)+
  geom_point(data = all_data, 
             aes(x = generation, y = predicted_freq, group = pop, color = env),
             alpha = 0.2)+
  geom_line(data = mean_data, 
            aes(x = generation, y = mean_freq, group = pop, color = env),
            alpha = 0.5, linewidth = 1.5) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("rec-1 mutant frequency")

```


# selection coefficient


## logit

```{r}
logit_data <- all_data %>%
  mutate(logit_freq = qlogis(predicted_freq))
logit_mean <- mean_data %>%
  mutate(logit_mean_freq = qlogis(mean_freq))

ggplot() +
  geom_point(data = logit_data, 
             aes(x = generation, y = logit_freq, group = pop, color = env),
             alpha = 0.2)+
  geom_line(data = logit_mean, 
            aes(x = generation, y = logit_mean_freq, group = pop, color = env),
            alpha = 0.4, linewidth = 1.5) +
  theme_classic()+
  theme(
  axis.title = element_text(size = 15),
  axis.text = element_text(size = 12))+
  xlab("Generation") +
  ylab("logit rec-1 mutant frequency")
```


## check data distribution

```{r}
hist(x = mean_data$mean_freq, breaks = 30)
hist(x = logit_mean$logit_mean_freq, breaks = 30)
```

## fit model

### fit complexe

fixed effect : intercept, generation, env, genetic background

random effect (on intercept and slope): pop

-> boundary (singular) fit: see help('isSingular')

-> meaning that model is too complexe (random effect structure)

-> corr for pop btw intercept and generation is -1

```{r}
fit_1 <- lmer(
  logit_mean_freq ~ 1+ generation*env*comp_back + (generation | pop ),
  data = logit_mean
)

summary(fit_1)
```

### fit simplified - good

fixed effect : intercept, generation, env, genetic background

random effect (on intercept): pop


```{r}
fit_2 <- lmer(
  logit_mean_freq ~ 1+ generation*env*comp_back + (1| pop ),
  data = logit_mean
)

summary(fit_2)
```
### fit batch effect

Anova P value >> 0.05 --> no significant influence of batch effect

```{r}
fit_batch <- lmer(
  logit_mean_freq ~ 1+ generation*env*comp_back + (1 | pop ) +(1 | experiment_batch),
  data = logit_mean
)

summary(fit_batch)
anova(fit_2, fit_batch)
```


### slope

```{r}
library(emmeans)

# extract generation effect (slope) for every pair of env*background
slopes <- emtrends(
  fit_2,
  ~ env * comp_back,
  var = "generation"
)

selection_table <- summary(slopes, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )

selection_table

pairs(slopes, by = "comp_back")
pairs(slopes, by = "env")
```



# Visualization

## selection coef

```{r, fig.width=10, fig.height=5}
ggplot(selection_table, aes(x = comp_back, y = selection_effect, color = env)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  theme_classic() +
  xlab("Genetic background") +
  ylab("Selection coefficient") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```


# Comparaison

## anova

```{r}
library(lmerTest)

anova(fit_2, type = 3)

# compare genetic backgrounds
background_contrasts <- pairs(
  slopes,
  by = "env",
  adjust = "tukey"
)

background_contrasts

# compare environments
environment_contrasts <- pairs(
  slopes,
  by = "comp_back",
  adjust = "tukey"
)

environment_contrasts
```

```{r}
p_background_within_env <- ggplot(
  selection_table,
  aes(x = comp_back, y = selection_effect, colour = env)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.6,
    colour = "grey45"
  ) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    width = 0.12,
    linewidth = 0.8
  ) +
  geom_point(
    size = 3.2,
    shape = 16
  ) +
  facet_wrap(~ env, nrow = 1) +
  labs(
    x = "Genetic background",
    y = "Selection coefficient",
    title = "Selection coefficients across genetic backgrounds"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 15, face = "bold", colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.background = element_rect(fill = "grey90", colour = "black"),
    strip.text = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.4),
    panel.border = element_rect(colour = "black", linewidth = 0.7)
  )

p_background_within_env
```



```{r}
contrast_background <- pairs(
  slopes,
  by = "env",
  adjust = "tukey"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame()

p_contrast_background <- ggplot(
  contrast_background,
  aes(x = contrast, y = estimate)
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15,
    linewidth = 0.7
  ) +
  facet_wrap(~ env, scales = "free_x") +
  coord_flip() +
  theme_bw() +
  xlab("Genetic background comparison") +
  ylab("Difference in selection coefficient") +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 11, colour = "black"),
    strip.text = element_text(size = 13, face = "bold")
  )

p_contrast_background
```


```{r}


# env coef
slope_env <- emtrends(
  fit_2,
  ~ env,
  var = "generation",
  weights = "equal"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  transmute(
    group_type = "Environment",
    group = as.character(env),
    selection_effect = generation.trend,
    SE = SE,
    df = df,
    lower_CI = lower.CL,
    upper_CI = upper.CL,
    p.value = p.value
  )

# genetic background coef
slope_background <- emtrends(
  fit_2,
  ~ comp_back,
  var = "generation",
  weights = "equal"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  transmute(
    group_type = "Genetic background",
    group = as.character(comp_back),
    selection_effect = generation.trend,
    SE = SE,
    df = df,
    lower_CI = lower.CL,
    upper_CI = upper.CL,
    p.value = p.value
  )

# summary table
summary_selection <- bind_rows(
  slope_env,
  slope_background
) %>%
  mutate(
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

summary_selection

```





```{r}
x_levels <- summary_selection %>%
  arrange(factor(group_type, levels = c("Environment", "Genetic background"))) %>%
  pull(group)

summary_selection <- summary_selection %>%
  mutate(
    group = factor(group, levels = x_levels)
  )

y_range <- diff(
  range(
    c(summary_selection$lower_CI, summary_selection$upper_CI),
    na.rm = TRUE
  )
)

summary_selection <- summary_selection %>%
  mutate(
    label_y = upper_CI + 0.08 * y_range
  )

group_cols <- c(
  "NGM"    = "grey50",
  "Zinc"   = "black",
  "ECA191" = "#6eabec",
  "ECA722" = "#4e5bcc",
  "ECA723" = "#6834d7",
  "ECA730" = "#b734d7"
)
ggplot(summary_selection,aes(x = group, y = selection_effect)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.6,colour = "grey45") +
  geom_errorbar(aes( ymin = lower_CI, ymax = upper_CI, colour = group),
                width = 0.12, linewidth = 0.9) +
  geom_point( aes( fill = group, colour = group),
              size = 4,shape = 21,stroke = 0.9) +
  scale_fill_manual(values = group_cols) +
  scale_colour_manual(values = group_cols) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.20))) +
  labs(x = NULL, y = "Selection coefficient",
       title = "Overall selection coefficients by environment and genetic background") +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.title.y = element_text( size = 15, face = "bold", colour = "black" ),
        axis.text = element_text( size = 12, colour = "black"), 
        axis.text.x = element_text( angle = 45, hjust = 1, vjust = 1 ),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(colour = "grey85",linewidth = 0.4),
        panel.border = element_rect( colour = "black", linewidth = 0.7))

```

# take out generation 0

```{r}
logit_mean_0 <- logit_mean%>%
  filter(generation != 0)
```


## fit model
fixed effect : intercept, generation, envi, genetic background

random effect (on intercept): pop


```{r}
fit_0 <- lmer(
  logit_mean_freq ~ 1+ generation*env*comp_back + (1| pop ),
  data = logit_mean_0
)

summary(fit_0)
```

### slope

```{r}
library(emmeans)

# extract generation effect (slope) for every pair of envi*background
slopes <- emtrends(
  fit_2,
  ~ env * comp_back,
  var = "generation"
)

selection_table <- summary(slopes, infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  dplyr::rename(
    selection_effect = generation.trend,
    SE = SE,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )

selection_table

pairs(slopes, by = "comp_back")
pairs(slopes, by = "env")
```



## Visualization

```{r, fig.width=10, fig.height=5}
ggplot(selection_table, aes(x = comp_back, y = selection_effect, color = env)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  theme_classic() +
  xlab("Genetic background") +
  ylab("Selection coefficient") +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

```{r}
p_background_within_env <- ggplot(
  selection_table,
  aes(x = comp_back, y = selection_effect, colour = env)
) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.6,
    colour = "grey45"
  ) +
  geom_errorbar(
    aes(ymin = lower_CI, ymax = upper_CI),
    width = 0.12,
    linewidth = 0.8
  ) +
  geom_point(
    size = 3.2,
    shape = 16
  ) +
  facet_wrap(~ env, nrow = 1) +
  labs(
    x = "Genetic background",
    y = "Selection coefficient",
    title = "Selection coefficients across genetic backgrounds"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 15, face = "bold", colour = "black"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.background = element_rect(fill = "grey90", colour = "black"),
    strip.text = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.4),
    panel.border = element_rect(colour = "black", linewidth = 0.7)
  )

p_background_within_env
```



```{r}
contrast_background <- pairs(
  slopes,
  by = "env",
  adjust = "tukey"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame()

p_contrast_background <- ggplot(
  contrast_background,
  aes(x = contrast, y = estimate)
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15,
    linewidth = 0.7
  ) +
  facet_wrap(~ env, scales = "free_x") +
  coord_flip() +
  theme_bw() +
  xlab("Genetic background comparison") +
  ylab("Difference in selection coefficient") +
  theme(
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 11, colour = "black"),
    strip.text = element_text(size = 13, face = "bold")
  )

p_contrast_background
```


```{r}

# env coef
slope_env <- emtrends(
  fit_2,
  ~ env,
  var = "generation",
  weights = "equal"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  transmute(
    group_type = "envronment",
    group = as.character(env),
    selection_effect = generation.trend,
    SE = SE,
    df = df,
    lower_CI = lower.CL,
    upper_CI = upper.CL,
    p.value = p.value
  )

# genetic background coef
slope_background <- emtrends(
  fit_2,
  ~ comp_back,
  var = "generation",
  weights = "equal"
) %>%
  summary(infer = c(TRUE, TRUE)) %>%
  as.data.frame() %>%
  transmute(
    group_type = "Genetic background",
    group = as.character(comp_back),
    selection_effect = generation.trend,
    SE = SE,
    df = df,
    lower_CI = lower.CL,
    upper_CI = upper.CL,
    p.value = p.value
  )

# summary table
summary_selection <- bind_rows(
  slope_env,
  slope_background
) %>%
  mutate(
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

summary_selection

```


```{r}
x_levels <- summary_selection %>%
  arrange(factor(group_type, levels = c("Environment", "Genetic background"))) %>%
  pull(group)

summary_selection <- summary_selection %>%
  mutate(
    group = factor(group, levels = x_levels)
  )

y_range <- diff(
  range(
    c(summary_selection$lower_CI, summary_selection$upper_CI),
    na.rm = TRUE
  )
)

summary_selection <- summary_selection %>%
  mutate(
    label_y = upper_CI + 0.08 * y_range
  )

group_cols <- c(
  "NGM"    = "grey50",
  "ZINC"   = "black",
  "ECA191" = "#6eabec",
  "ECA722" = "#4e5bcc",
  "ECA723" = "#6834d7",
  "ECA730" = "#b734d7"
)
ggplot(summary_selection,aes(x = group, y = selection_effect)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.6,colour = "grey45") +
  geom_errorbar(aes( ymin = lower_CI, ymax = upper_CI, colour = group),
                width = 0.12, linewidth = 0.9) +
  geom_point( aes( fill = group, colour = group),
              size = 4,shape = 21,stroke = 0.9) +
  geom_text( aes(y = label_y, label = sig_label),size = 4.5,
             fontface = "bold",colour = "black") +
  scale_fill_manual(values = group_cols) +
  scale_colour_manual(values = group_cols) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.20))) +
  labs(x = NULL, y = "Selection coefficient",
       title = "Overall selection coefficients by environment and genetic background") +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.title.y = element_text( size = 15, face = "bold", colour = "black" ),
        axis.text = element_text( size = 12, colour = "black"), 
        axis.text.x = element_text( angle = 45, hjust = 1, vjust = 1 ),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(colour = "grey85",linewidth = 0.4),
        panel.border = element_rect( colour = "black", linewidth = 0.7))

```
